---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  # print 15 lines of table output
  tidy.opts = list(width.cutoff = 60
               , max.print = 15
  )
)
```

# journalR

<!-- badges: start -->
<!-- badges: end -->

The goal of journalR is to ease real-world scientific publication writing. 

## Installation

You can install the development version of journalR from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("epi-sam/journalR")
```

## Basics

In the Platonic ideal world, a journal is written in pure LaTeX, Rmarkdown, or
the like.

In the real world, journals are written collaboratively in Word, with messy 
versioning, multiple copies of the same file, broken citations, and enjoy 
constant updates and handoffs between team members.

Sometimes, you just need to copy and paste some numbers.

Enter journalR.

```{r example}
library(journalR)
# data.table is not required, but the author finds it helpful.
library(data.table)
```

Let's say you have done some statistical analysis.

```{r mtcars}
DT <- data.table::as.data.table(mtcars)

# calculate mean/lower/upper 95% UI for mpg by cyl
DT_mpg <- DT[
   , .(
        mean  = mean(mpg)
      , lower = quantile(mpg, 0.025)
      , upper = quantile(mpg, 0.975)
   )
   , by = cyl
]

print(DT_mpg)
```

Once they're ready for presenation, you'll want to format them.

Enter `format_journal_df()`:

- Format a mean/lower/upper set of three columns into a `mean (lower -- upper)` string.
   - Ready for easy copy/paste into Word or similar.
- The user must provide a data type (`d_type`), which tells the function how to format the numbers.
   - mean/lower/upper relationships are asserted (lower < mean < upper) before formatting.

```{r format}
DT_mpg |>
   journalR::format_journal_df(
      d_type = "count"
   )
```

## Options 

The user's data.frame may be formatted differently than the simple case above, with different data types.

- Things get messy in the real world!

```{r options}
# build a mean/lower/uppper summary table with some percentage from DT
DT_summary <- DT[
   , .(
        n            = .N
      , mean_mpg     = mean(mpg)
      , lower_mpg    = quantile(mpg, 0.025)
      , upper_mpg    = quantile(mpg, 0.975)
      , pct_vshape   = mean(vs) 
      , lower_vshape = quantile(vs, 0.025) 
      , upper_vshape = quantile(vs, 0.975) 
   )
   , by = cyl
]

```

As long as your table is tidy, it can be formatted.

```{r options_formatted}
DT_summary |>
   journalR::format_journal_df(
      d_type               = c("prop")
      , new_var            = "pct_vshape_fmt" # specify output colname (must be a new one)
      , central_var        = "pct_vshape" 
      , lower_var          = "lower_vshape"
      , upper_var          = "upper_vshape"
      , remove_clu_columns = FALSE # retain the original columns if desired
   )
```

## Styles

Two formatting styles are built into journalR, and the user may extend this.

- Nature (defautl)
- Lancet (particular)

```{r style_lancet, include=FALSE}
lancet <- data.table::as.data.table(stack(style_lancet()))
data.table::setcolorder(lancet, "ind")
data.table::setnames(lancet, c("ind", "values"), c("key", "value"))
```

Styles are a key/value list of metadata controlling rounding and formatting.

Here is an example of a very particular style, and how these values affect presentation tables. 

```{r style_lancet_show}
# print(style_lancet()) # as a list
print(lancet) # formatted a little nicer
```

Counts receive a non-standard thousands separator.

```{r lancet_format_counts}
journalR::format_journal_df(
   data.frame(
        mean          = c(55.8e3, 54.7e6)
      , lower         = c(50.7e3, 48.6e6)
      , upper         = c(60.7e3, 59.6e6)
   )
   , d_type = "count"
   , style = "lancet"
)
```

Netagives are handled gracefully.

- Lancet decimals are `mid_dot()` characters

```{r lancet_format_prop}
journalR::format_journal_df(
   data.table::data.table(
                  data_space    = c("all_positive", "mixed_negative", "all_negative")
                , mean          = c(.5584654, -0.15665, -0.1321684)
                , lower         = c(.5076231, -0.25321, -0.235321)
                , upper         = c(.6076589, 1.365432, -0.056549)
             )
   , d_type = "prop"
   , style = "lancet"
)
```

## Extending styles

Users may define their own styles by creating a named list with the same keys as the built-in styles.

```{r new_style}

journalR::new_style(
   style_name                 = 'my_style'
   , digits_round_prop        = 3
   , digits_sigfig_count      = 5
   , nsmall                   = 3
   , decimal.mark             = "."
   , negative_sign            = "-"
   , big.mark_count           = ","
   , mean_neg_text            = "Negative"
   , UI_only                  = FALSE
   , UI_text                  = "95%UI "
   , assert_clu_relationships = TRUE
   , is_lancet                = FALSE
   , allow_thousands          = FALSE
)
```


Perhaps the user prefers standard thousands separators for counts.

```{r apply_new_style_counts}
journalR::format_journal_df(
   data.frame(
        mean          = c(55.8e3, 54.7e6)
      , lower         = c(50.7e3, 48.6e6)
      , upper         = c(60.7e3, 59.6e6)
   )
   , d_type = "count"
   , style  = "my_style"
)
```

Or perhaps thousands should be treated like millions with high precision.

```{r new_style_2}
journalR::new_style(
   style_name                 = 'my_style2'
   , digits_round_prop        = 3
   , digits_sigfig_count      = 5
   , nsmall                   = 3
   , decimal.mark             = "."
   , negative_sign            = "-"
   , big.mark_count           = ","
   , mean_neg_text            = "Negative"
   , UI_only                  = FALSE
   , UI_text                  = "95%UI "
   , assert_clu_relationships = TRUE
   , is_lancet                = FALSE
   , allow_thousands          = TRUE
)
journalR::format_journal_df(
   data.frame(
        mean          = c(55832, 54.7e6)
      , lower         = c(50797, 48.6e6)
      , upper         = c(60754, 59.6e6)
   )
   , d_type = "count"
   , style  = "my_style2"
)
```


Perhaps the user's values have higher precision - you can allow for that with your custom style.

```{r apply_new_style_prop}
journalR::format_journal_df(
   data.table::data.table(
                  data_space    = c("all_positive", "mixed_negative", "all_negative")
                , mean          = c(.5584654, -0.15665, -0.1321684)
                , lower         = c(.5076231, -0.25321, -0.235321)
                , upper         = c(.6076589, 1.365432, -0.056549)
             )
   , d_type = "prop"
   , style = "my_style"
)
```

Done. 
